---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# GPEDM

<!-- badges: start -->
<!-- badges: end -->

**Disclaimer: This is a very much a work in progress. Use at your own risk.**

This is version `r packageVersion("GPEDM")`

This package contains functions for fiting hierarchical, separable length scale GP models with automatic relevance determination (ARD) for use in Empirical Dynamic Modeling (EDM) and other applications. This is an adaptation of code originally developed by Stephan Munch in MATLAB.  

The main function is `fitGP` which is used to train the model and can also produce predictions if desired. Use `summary.GP` to view a summary, `predict.GP` to generate other or additional predictions from a fitted model, `plot.GPpred` to plot observed and predicted values and `getconditionals` to obtain conditional reponses. Also available for use are the functions `makelags` which can be used to create delay vectors, and `getR2` for getting R2 values. See the (not yet created) vignette for more detailed instructions.

## Installation

To install the package:

``` r
install.packages("devtools") #if required
devtools::install_github("tanyalrogers/GPEDM")
```
If you are using Windows, you may need to install [Rtools](https://cran.r-project.org/bin/windows/Rtools/), which allows you to complile the C++ code.

## Basic Use

Here are some simulated data from 2 populations with theta logistic dynamics. The data contain some small lognormal process noise, and the populations have different theta values.

```{r data}
library(GPEDM)
data("thetalog2pop")

pA=subset(thetalog2pop,Population=="PopA")
pB=subset(thetalog2pop,Population=="PopB")
N=nrow(pA)
par(mfrow=c(2,2),mar=c(4,4,2,1))
plot(Abundance~Time,data=pA,type="l",main="PopA")
plot(Abundance~Time,data=pB,type="l",main="PopB")
plot(pA$Abundance[1:(N-1)],pA$Abundance[2:N],
     xlab="Abundance t",ylab="Abundance t+1",main="PopA")
plot(pB$Abundance[1:(N-1)],pB$Abundance[2:N],
     xlab="Abundance t",ylab="Abundance t+1",main="PopB")
```

Here is how you might set up a hierarchical time-delay embedding model. In this example, `Abundance` is the response variable (`y`). We will use an embedding dimension (`E`) of 3 and time delay (`tau`) of 1. `Population` indicates which population the data are from, so it is included under `pop`. Since the data are on somewhat different scales and don't necessarily represent the same 'units', we will use local (within population) data scaling, as opposed to global. Just for fun, we will also request leave-one-out predictions.   

```{r example}
tlogtest=fitGP(data = thetalog2pop, y = "Abundance", pop = "Population", E=3, tau=1, 
               scaling = "local", predictmethod = "loo")
summary(tlogtest)
```

From the summary, we can see that ARD has (unsurprisingly) deemed lags 2 and 3 to be unimportant (length scales are 0), so E=1 is probably sufficient. The dynamic correlation (rho) tells us the degree to which the dynamics are correlated (they are are rather dissimilar in this case). Since the simulated data don't contain that much noise, the R-squared values are quite high.  

### Plot observed and predicted values

The observed and predicted values can be found under `model$insampresults` or `model$outsampresults`.

We can make a quick plot of the observed and predicted values using `plot`. Standard error bands are included in the time series plots, although they're a little hard to see in this example.

```{r plot1}
plot(tlogtest)
```

To get the in-sample predictions:

```{r plot2}
plot(tlogtest, plotinsamp = T)
```

If you prefer ggplot:

```{r ggplot-timeseries, fig.height=2.5}
library(ggplot2)
ggplot(tlogtest$insampresults,aes(x=timestep,y=predmean)) +
  facet_wrap(pop~., scales = "free") +
  geom_line() + geom_ribbon(aes(ymin=predmean-predfsd,ymax=predmean+predfsd), alpha=0.4) +
  geom_point(aes(y=obs)) +
  theme_bw()
```

### Plot conditional responses

The function `getconditionals` will compute and plot conditional responses to each input variable (other input varibles set to their mean value). From this we can also clearly see that lags 2 and 3 have no impact, and we can see how the lag 1 dynamics of the 2 populations differ.

```{r conditionals}
con=getconditionals(tlogtest)
```

If you prefer ggplot:

```{r ggplot-conditionals}
#have to convert conditionals output to long format
#there may be a more concise way to do this
library(tidyr)
npreds=length(grep("_yMean",colnames(con)))
conlong1=gather(con[,1:(npreds+1)],x,xValue,2:(npreds+1))
conlong2=gather(con[,c(1,(npreds+2):(2*npreds+1))],ym,yMean,2:(npreds+1))
conlong3=gather(con[,c(1,(2*npreds+2):(3*npreds+1))],ys,ySD,2:(npreds+1))
conlong=cbind.data.frame(conlong1,yMean=conlong2$yMean,ySD=conlong3$ySD)
ggplot(conlong,aes(x=xValue,y=yMean)) +
  facet_grid(pop~x, scales = "free") +
  geom_line() + geom_ribbon(aes(ymin=yMean-ySD,ymax=yMean+ySD), alpha=0.4) +
  theme_bw()
```

### Plot inverse length scales

The model hyperparameters are located under `model$pars`. If you have $n$ predictors, the first $n$ values of `pars` will be the length scales. Note that if you use `E` and `tau`, the names of the predictors in the input data frame will be stored under `model$inputs$x_names`, and the names of the lagged predictors corresponding to the inverse length scales will be stored under `model$inputs$x_names2`.

```{r ls, fig.height=3}
predvars=tlogtest$inputs$x_names2
npreds=length(predvars)
lscales=tlogtest$pars[1:npreds]
par(mar=c(4,4,1,1))
plot(factor(predvars),lscales,xlab="Predictor",ylab="Inverse length scale")
```

### Other types of predictions

We can use the `predict` function to get various types of predictions from a fitted model. Leave-one-out, `predictmethod = "loo"` is one option (above, we got these predictions at the same time we fit the model, but this is not necessary). The following obtains sequential (leave-future-out) predictions using the training data. 

```{r predict}
#sequential predictions (they should improve over time)
seqpred=predict(tlogtest,predictmethod = "sequential")
plot(seqpred)
```

You could, alternatively, supply new data for which to make predictions. In that case, you would supply a new data frame (`newdata`). In our example, this data frame which should contain columns `Abundance` and `Population`.  

A common approach when fitting these models is to split the available data into a training and test dataset. For instance, say we wanted a single-population model for PopA with 2 time lags, and we wanted to use the first 40 data points as training data, and the last 10 points as test data. For that we could do the following.

```{r predict2, fig.height=3}
pAtrain=pA[1:40,]
pAtest=pA[41:50,]
tlogtest2=fitGP(data = pAtrain, y = "Abundance", E=2, tau=1,
                newdata = pAtest)
plot(tlogtest2)
```

If you don't want the first `E*tau` points getting excluded from your test data, generate the lags beforehand, then split the data (don't use `E` and `tau` options in `fitGP`). See **Specifying training data** (option 1).

```{r predict3, fig.height=3}
pAlags=makelags(pA, y = "Abundance", E=2, tau=1)
pAdata=cbind(pA,pAlags)
pAtrain=pAdata[1:40,]
pAtest=pAdata[41:50,]
tlogtest3=fitGP(data = pAtrain, y = "Abundance", x=colnames(pAlags),
                newdata = pAtest)
plot(tlogtest3)
```


### Making forecasts

You can construct a forecast matrix using `makelags` by setting `forecast=T` and supplying this as `newdata`. In order to do this (as the code is currently written), you have to generate the lags beforehand for *both* the training data and the forecast (you cannot use `E` and `tau` in `fitGP`, you have to use `makelags` for both, and the settings in `makelags` should match, other than `forecast`, See **Specifying training data** option 1). It is a good idea to include the `time` argument when doing this.

```{r forecast, fig.height=3}
lags1=makelags(thetalog2pop,y=c("Abundance"),pop="Population",time="Time",E=3,tau=1)
fore1=makelags(thetalog2pop,y=c("Abundance"),pop="Population",time="Time",E=3,tau=1,forecast = T)
data1=cbind(thetalog2pop, lags1)

tlogfore=fitGP(data = data1, y = "Abundance", x=c("Abundance_1","Abundance_2","Abundance_3"), 
               pop = "Population", time = "Time", scaling = "local", newdata = fore1)

ggplot(tlogfore$insampresults,aes(x=timestep,y=predmean)) +
  facet_wrap(pop~., scales = "free") +
  geom_line() + geom_ribbon(aes(ymin=predmean-predsd,ymax=predmean+predsd), alpha=0.4) +
  geom_point(aes(y=obs)) +
  geom_point(data=tlogfore$outsampresults, aes(y=predmean), color="red") +
  geom_errorbar(data=tlogfore$outsampresults,
                aes(ymin=predmean-predsd,ymax=predmean+predsd),color="red") +
  theme_bw()
```


## Specifying training data

There are several ways that the training data for a model can be specified. 

A. supply data frame `data`, plus column names or indices for `y` and `x`.  
B. supply a vector for `y` and a vector or matrix for `x`.  

For each of the above 2 options, there are 3 options for specifying the predictors.  

1. supplying `y` and `x` (omitting `E` and `tau`) will use the columns of `x` as predictors. This allows for the most customization.  
2. supplying `y`, `E`, and `tau` (omitting `x`) will use `E` lags of `y` (with spacing `tau`) as predictors. This is equivalent to option 3 with `x`=`y`.  
3. supplying `y`, `x`, `E`, and `tau` will use `E` lags of *each column* of `x` (with spacing `tau`) as predictors. Do not use this option if `x` already contains lags, in that case use option 1.  

The above example is method A2. Method A1/B1 allows for the most customization of response and predictor variables including mixed embeddings. 

Options 2 and 3 exist for convenience, but for the most control over the model and predictions, it is best to use option 1: use `makelags()` to generate any lags beforehand and pass appropriate columns to `fitGP`, rather than rely on `fitGP` to generate lags internally. Option A will make more sense if your data are already in a data frame, option B may make more sense if you are doing simulations and just have a bunch of vectors and matrices.

The `pop` argument is optional in all of the above cases. If omitted, a single population is assumed.

```{r other_methods}
set.seed(10)
thetalog2pop$othervar=rnorm(nrow(thetalog2pop))
yvec=thetalog2pop$Abundance
popvec=thetalog2pop$Population
#function 'makelags' can be used to generate a lag matrix
#be sure to include 'pop' if data contain multiple pops to prevent crossover
xmat=makelags(y=thetalog2pop[,c("Abundance","othervar")],pop=popvec,E=2,tau=1)
thetalog2pop2=cbind(thetalog2pop,xmat)

#Method A1
ma1=fitGP(data=thetalog2pop2,y="Abundance",x=c("Abundance_1","othervar"),
         pop="Population",scaling="local")

#Method B1
#like A1, but your data aren't in a data frame
mb1=fitGP(y=yvec,x=xmat,pop=popvec,scaling="local")

#Method B2
#like A2, but your data aren't in a data frame
mb2=fitGP(y=yvec,pop=popvec,E=2,tau=1,scaling="local")

#Method A3
#generate lags of multiple predictors internally
ma3=fitGP(data=thetalog2pop2,y="Abundance",x=c("Abundance","othervar"),
         pop="Population",E=2,tau=1,scaling="local")


summary(ma1)

summary(mb1)

summary(mb2)

summary(ma3)
```

## Variable Timestep Method (for missing data)

By default, `fitGP` excludes any rows that contain missing values (NAs) in either the response or predictor variables. Thus, in a delay embedding model, a missing datapoint in the middle of a time series will be excluded as will the following `E` values (assuming `tau` is 1). Stephan Munch and Bethany Johnson developed a method whereby using the time spacing between a value and its lagged predictors as another predictor, we can exclude less data, generate predictions for all non-missing timepoints, and by adjusting the spacing in the forecast matrix, generate forecasts multiple timesteps into the future. The forthcoming paper will contain more details.  

Using this method requires that lag predictors be generated beforehand (option 1 in **Specifying training data**). The `makelags` function has an option for the variable timestep method (`vtimestep=TRUE`) that will generate the time difference lags automatically. Including the `time` argument is strongly recommended, and you definitely need to include it if the rows with missing values have already been removed and/or the timesteps are uneven. Argument `Tdiff_max` can be used to set the max time difference value considered (e.g. if you have one large time gap that you don't want to use this method for).

```{r vtimestep, fig.height=3}
#just using one population for now, it will also work with multiple populations

#add some missing values
pAmiss=pA
pAmiss[c(7,10,15,23,30,40),"Abundance"]=NA

#standard method
pAmisslags=makelags(data=pAmiss, y="Abundance", time="Time", E=2, tau=1)
head(cbind(pAmiss,pAmisslags),15)

#variable timestep method
pAmisslags=makelags(data=pAmiss, y="Abundance", time="Time", E=2, tau=1, vtimestep=T)
head(cbind(pAmiss,pAmisslags),15)

pAmissdata=cbind(pAmiss,pAmisslags)
vtdemo=fitGP(data=pAmissdata, y="Abundance", x=colnames(pAmisslags), time="Time")

summary(vtdemo)

ggplot(vtdemo$insampresults,aes(x=timestep,y=predmean)) +
  geom_line() + 
  geom_ribbon(aes(ymin=predmean-predsd,ymax=predmean+predsd), alpha=0.4, color="black") +
  geom_point(aes(y=obs)) +
  theme_bw()
```

You can also generate a forecast matrix using the variable timestep method. The number of timesteps to forecast can be specified with `Tdiff_fore`.

```{r vtimestep2, fig.height=3}
pAmissfore=makelags(data=pAmiss, y="Abundance", time="Time", E=2, tau=1,vtimestep=T,  
                    forecast=T, Tdiff_fore=c(1,2))
pAmissfore

vtpred=predict(vtdemo, newdata = pAmissfore)

ggplot(vtdemo$insampresults,aes(x=timestep,y=predmean)) +
  geom_line() + 
  geom_ribbon(aes(ymin=predmean-predsd,ymax=predmean+predsd), alpha=0.4, color="black") +
  geom_point(aes(y=obs)) +
  geom_point(data=vtpred$outsampresults, aes(y=predmean), color="red") +
  geom_errorbar(data=vtpred$outsampresults,
                aes(ymin=predmean-predsd,ymax=predmean+predsd),color="red") +
  theme_bw()
```

### Using augmentation data

When using the variable timestep method, the function `makelags` can also be used to generate an augmentation data matrix that can be passed to `fitGP`. This should work as long as the settings in `makelags` match, except use `augment=TRUE`. When the augmentation table is generated, `makelags` will print a table showing the original number of each Tdiff combination in the original dataset (Freq), and the total number of with the augmentation data included (Freq_new). Combinations are added up to `nreps`, if possible (currently defaults to 10). By default, only Tdiff combinations that appear in the original dataset are used, however, if you supply a vector `Tdiff_fore`, then the augmentation matrix will include or all possible combinations of the Tdiff values supplied in `Tdiff_fore`.

```{r aug, fig.height=3}
pAmisslags=makelags(data=pAmiss, y="Abundance", time="Time", E=2, tau=1, vtimestep=T)
pAaug=makelags(data=pAmiss, y="Abundance", time="Time", E=2, tau=1, vtimestep=T,  
               augment=T)
pAaug

pAmissdata=cbind(pAmiss,pAmisslags)
vtdemo_aug=fitGP(data=pAmissdata, y="Abundance", x=colnames(pAmisslags), time="Time",
                 augdata=pAaug,newdata=pAmissfore)
summary(vtdemo_aug)

ggplot(vtdemo_aug$insampresults,aes(x=timestep,y=predmean)) +
  geom_line() + 
  geom_ribbon(aes(ymin=predmean-predsd,ymax=predmean+predsd), alpha=0.4, color="black") +
  geom_point(aes(y=obs)) +
  geom_point(data=vtdemo_aug$outsampresults, aes(y=predmean), color="red") +
  geom_errorbar(data=vtdemo_aug$outsampresults,
                aes(ymin=predmean-predsd,ymax=predmean+predsd),color="red") +
  theme_bw()

vtaloo=predict(vtdemo_aug,predictmethod = "loo")
vtaseq=predict(vtdemo_aug,predictmethod = "sequential")
vtloo=predict(vtdemo,predictmethod = "loo")
vtseq=predict(vtdemo,predictmethod = "sequential")
vtanew=predict(vtdemo_aug,newdata=pAmissfore)
```


## References

Munch, S. B., Poynor, V., and Arriaza, J. L. 2017. Circumventing structural uncertainty: a Bayesian perspective on nonlinear forecasting for ecology. Ecological Complexity, 32:134.

Johnson, B., and Munch, S. B. 2022. An empirical dynamic modeling framework for missing or irregular samples. Ecological Modelling, 468:109948.


*Any advice on improving this package is appreciated.*

## Github Disclaimer

_This repository is a scientific product and is not official communication of the National Oceanic and Atmospheric Administration, or the United States Department of Commerce. All NOAA GitHub project code is provided on an ‘as is’ basis and the user assumes responsibility for its use. Any claims against the Department of Commerce or Department of Commerce bureaus stemming from the use of this GitHub project will be governed by all applicable Federal law. Any reference to specific commercial products, processes, or services by service mark, trademark, manufacturer, or otherwise, does not constitute or imply their endorsement, recommendation or favoring by the Department of Commerce. The Department of Commerce seal and logo, or the seal and logo of a DOC bureau, shall not be used in any manner to imply endorsement of any commercial product or activity by DOC or the United States Government._
