---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# GPEDM

<!-- badges: start -->
<!-- badges: end -->

**Disclaimer: This is a very much a work in progress. Use at your own risk.**

This package contains functions for fiting hierarchical, separable length scale GP models with automatic relevance determination (ARD) for use in Empirical Dynamic Modeling (EDM) and other applications. This is an adaptation of code originally developed by Stephan Munch in MATLAB.  

The main function is `fitGP` which is used to train the model and can also produce predictions if desired. Use `summary.GP` to view a summary, `predict.GP` to generate other or additional predictions from a fitted model, `plot.GPpred` to plot observed and predicted values and `getconditionals` to obtain conditional reponses. Also available for use are the functions `makelags` which can be used to create delay vectors, and `getR2` for getting R2 values. See the (not yet created) vignette for more detailed instructions.

## Installation

To install the package:

``` r
install.packages("devtools") #if required
devtools::install_github("tanyalrogers/GPEDM")
```
If you are using Windows, you may need to install [Rtools](https://cran.r-project.org/bin/windows/Rtools/), which allows you to complile the C++ code.

## Example

Here are some simulated data from 2 populations with theta logistic dynamics. The data contain some small lognormal process noise, and the populations have different theta values.

```{r data}
library(GPEDM)
data("thetalog2pop")

pA=subset(thetalog2pop,Population=="PopA")
pB=subset(thetalog2pop,Population=="PopB")
N=nrow(pA)
par(mfrow=c(2,2),mar=c(4,4,2,1))
plot(Abundance~Time,data=pA,type="l",main="PopA")
plot(Abundance~Time,data=pB,type="l",main="PopB")
plot(pA$Abundance[1:(N-1)],pA$Abundance[2:N],
     xlab="Abundance t",ylab="Abundance t+1",main="PopA")
plot(pB$Abundance[1:(N-1)],pB$Abundance[2:N],
     xlab="Abundance t",ylab="Abundance t+1",main="PopB")
```

Here is how you might set up a hierarchical time-delay embedding model. In this example, `Abundance` is the response variable (`yd`). We will use an embedding dimension (`E`) of 3 and time delay (`tau`) of 1. `Population` indicates which population the data are from, so it is included under `pop`. Since the data are on somewhat different scales and don't necessarily represent the same 'units', we will use local (within population) data scaling, as opposed to global. Just for fun, we will also request leave-one-out predictions.   

```{r example}
tlogtest=fitGP(data = thetalog2pop, yd = "Abundance", pop = "Population", E=3, tau=1, 
               scaling = "local", predictmethod = "loo")
summary(tlogtest)
```

From the summary, we can see that ARD has (unsurprisingly) deemed lags 2 and 3 to be unimportant (length scales are 0), so E=1 is probably sufficient. The dynamic correlation (rho) tells us the degree to which the dynamics are correlated (they are are rather dissimilar in this case). Since the simulated data don't contain that much noise, the R-squared values are quite high.  

We can examine the observed and predicted values using `plot`. Standard error bands are included in the time series plots, although they're a little hard to see in this example.

```{r plot1}
plot(tlogtest)
```

To get the in-sample predictions:

```{r plot2}
plot(tlogtest, plotinsamp = T)
```

If you prefer ggplot:

```{r ggplot timeseries}
library(ggplot2)
ggplot(tlogtest$insampresults,aes(x=timestep,y=predmean)) +
  facet_wrap(pop~., scales = "free") +
  geom_line() + geom_ribbon(aes(ymin=predmean-predfsd,ymax=predmean+predfsd), alpha=0.4) +
  geom_point(aes(y=obs)) +
  theme_bw()
```

The function `getconditionals` will compute and plot conditional responses to each input variable (other input varibles set to their mean value). From this we can also clearly see that lags 2 and 3 have no impact, and we can see how the lag 1 dynamics of the 2 populations differ.

```{r conditionals}
con=getconditionals(tlogtest)
```

If you prefer ggplot:

```{r ggplot conditionals}
#have to convert conditionals output to long format
#there may be a more concise way to do this
library(tidyr)
npreds=length(grep("_yMean",colnames(con)))
conlong1=gather(con[,1:(npreds+1)],x,xValue,2:(npreds+1))
conlong2=gather(con[,c(1,(npreds+2):(2*npreds+1))],ym,yMean,2:(npreds+1))
conlong3=gather(con[,c(1,(2*npreds+2):(3*npreds+1))],ys,ySD,2:(npreds+1))
conlong=cbind.data.frame(conlong1,yMean=conlong2$yMean,ySD=conlong3$ySD)
ggplot(conlong,aes(x=xValue,y=yMean)) +
  facet_grid(pop~x, scales = "free") +
  geom_line() + geom_ribbon(aes(ymin=yMean-ySD,ymax=yMean+ySD), alpha=0.4) +
  theme_bw()
```

Make a plot of the inverse length scale parameters. Note that if you use `E` and `tau`, the names of the predictors in the input data frame will be stored under `xd_names`, and the names of the lagged predictors corresponding to the inverse length scales will be stored under `xd_names2`.

```{r ls}
predvars=tlogtest$inputs$xd_names2
npreds=length(predvars)
lscales=tlogtest$pars[1:npreds]
plot(factor(predvars),lscales,xlab="Predictor",ylab="Inverse length scale")

```

We can use the `predict` function to get other types of predictions. The following obtains sequential predictions using the training data. 

```{r predict}
#sequential predictions (they should improve over time)
seqpred=predict(tlogtest,predictmethod = "sequential")
plot(seqpred)
```

You could, alternatively, supply new data for which to make predictions. In that case, you would supply a new data frame (`datanew`), which should contain columns `Abundance` and `Population`.  

A common approach when fitting these models is to split the available data into a training and test dataset. For instance, say we wanted a single-population model for PopA with 2 time lags, and we wanted to use the first 40 data points as training data, and the last 10 points as test data. For that we could do the following.

```{r predict2}
pAtrain=pA[1:40,]
pAtest=pA[41:50,]
tlogtest2=fitGP(data = pAtrain, yd = "Abundance", E=2, tau=1,
                datanew = pAtest)
plot(tlogtest2)
```

## Specifying training data

There are several ways that the training data for a model can be specified. 

A. supply data frame `data`, plus column names or indices for `yd` and `xd`.  
B. supply a vector for `yd` and a vector or matrix for `xd`.  

For each of the above 2 options, there are 3 options for specifying the predictors.  

1. supplying `yd` and `xd` (omitting `E` and `tau`) will use the columns of `xd` as predictors. This allows for the most customization.  
2. supplying `yd`, `E`, and `tau` (omitting `xd`) will use `E` lags of `yd` (with spacing `tau`) as predictors. This is equivalent to option 3 with `xd`=`yd`.  
3. supplying `yd`, `xd`, `E`, and `tau` will use `E` lags of *each column* of `xd` (with spacing `tau`) as predictors. Do not use this option if `xd` already contains lags, in that case use option 1.  

The above example is method A2. Method A1/B1 allows for the most customization of response and predictor variables including mixed embeddings.

The `pop` argument is optional in all of the above cases. If omitted, a single population is assumed.

```{r other methods}
set.seed(10)
thetalog2pop$othervar=rnorm(nrow(thetalog2pop))
yvec=thetalog2pop$Abundance
popvec=thetalog2pop$Population
#function 'makelags' can be used to generate a lag matrix
#be sure to include 'pop' if data contain multiple pops to prevent crossover
xmat=makelags(yd=thetalog2pop[,c("Abundance","othervar")],pop=popvec,E=2,tau=1)
thetalog2pop2=cbind(thetalog2pop,xmat)

#Method A1
ma1=fitGP(data=thetalog2pop2,yd="Abundance",xd=c("Abundance_1","othervar"),
         pop="Population",scaling="local")

#Method B1
#like A1, but your data aren't in a data frame
mb1=fitGP(yd=yvec,xd=xmat,pop=popvec,scaling="local")

#Method B2
#like A2, but your data aren't in a data frame
mb2=fitGP(yd=yvec,pop=popvec,E=2,tau=1,scaling="local")

#Method A3
#generate lags of multiple predictors internally
ma3=fitGP(data=thetalog2pop2,yd="Abundance",xd=c("Abundance","othervar"),
         pop="Population",E=2,tau=1,scaling="local")


summary(ma1)

summary(mb1)

summary(mb2)

summary(ma3)
```

## References

Munch, S. B., Poynor, V., and Arriaza, J. L. 2017. Circumventing structural uncertainty: a Bayesian perspective on nonlinear forecasting for ecology. Ecological Complexity, 32:134.


*Any advice on improving this package is appreciated.*
