---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# GPEDM

<!-- badges: start -->
<!-- badges: end -->

**Disclaimer: This is very much a work in progress. Use at your own risk and please report any problems.**

This is version `r packageVersion("GPEDM")`

This package contains functions for fiting hierarchical, separable length scale Gaussian process (GP) models with automatic relevance determination (ARD) for use in Empirical Dynamic Modeling (EDM) and other applications. This is an adaptation of code originally developed by Stephan Munch in MATLAB.  

The main function is `fitGP` which is used to train the model and can also produce predictions if desired. Use `summary` to view a summary, `predict` to generate other or additional predictions from a fitted model, `plot` to plot observed and predicted values, `update` to update an existing model, and `getconditionals` to obtain conditional reponses. Also available for use are the functions `makelags` which can be used to create delay vectors, and `getR2` for getting R2 values. See [Specifying training data] and the function documentation for more detailed instructions.

## Installation

To install the package:

``` r
install.packages("devtools") #if required
devtools::install_github("tanyalrogers/GPEDM")
```
If you are using Windows, you may need to install [Rtools](https://cran.r-project.org/bin/windows/Rtools/), which allows you to complile the C++ code.

## Basic Use

Here are some simulated data from 2 populations with theta logistic dynamics. The data contain some small lognormal process noise, and the populations have different theta values.

```{r data}
library(GPEDM)
data("thetalog2pop")

pA=subset(thetalog2pop,Population=="PopA")
pB=subset(thetalog2pop,Population=="PopB")
N=nrow(pA)
par(mfrow=c(2,2),mar=c(4,4,2,1))
plot(Abundance~Time,data=pA,type="l",main="PopA")
plot(Abundance~Time,data=pB,type="l",main="PopB")
plot(pA$Abundance[1:(N-1)],pA$Abundance[2:N],
     xlab="Abundance t",ylab="Abundance t+1",main="PopA")
plot(pB$Abundance[1:(N-1)],pB$Abundance[2:N],
     xlab="Abundance t",ylab="Abundance t+1",main="PopB")
```

Here is how you might set up a hierarchical time-delay embedding model. In this example, `Abundance` is the response variable (`y`). We will use an embedding dimension (`E`) of 3 and time delay (`tau`) of 1. `Population` indicates which population the data are from, so it is included under `pop`. Since the data are on somewhat different scales and don't necessarily represent the same 'units', we will use local (within population) data scaling, as opposed to global. Just for fun, we will also request leave-one-out predictions.   

```{r example}
tlogtest=fitGP(data = thetalog2pop, y = "Abundance", pop = "Population", E=3, tau=1, 
               scaling = "local", predictmethod = "loo")
summary(tlogtest)
```

From the summary, we can see that ARD has (unsurprisingly) deemed lags 2 and 3 to be unimportant (length scales are 0), so E=1 is probably sufficient. The dynamic correlation (rho) tells us the degree to which the dynamics are correlated (they are are rather dissimilar in this case). Since the simulated data don't contain that much noise, the R-squared values are quite high. For more information on choosing embedding parameters, see [this vignette](https://tanyalrogers.github.io/GPEDM/articles/choosingE.html). For more information on hierarchical modeling and the dynamic correlation, see [this other vignette](https://tanyalrogers.github.io/GPEDM/articles/hierarchical.html).  

### Plot observed and predicted values

The observed and predicted values can be found under `model$insampresults` or `model$outsampresults`.

We can make a quick plot of the observed and predicted values using `plot`. Standard error bands are included in the time series plots, although they're a little hard to see in this example.

```{r plot1}
plot(tlogtest)
```

To get the in-sample predictions:

```{r plot2}
plot(tlogtest, plotinsamp = T)
```

If you prefer ggplot:

```{r ggplot-timeseries, fig.height=2.5}
library(ggplot2)
ggplot(tlogtest$insampresults,aes(x=timestep,y=predmean)) +
  facet_wrap(pop~., scales = "free") +
  geom_line() + geom_ribbon(aes(ymin=predmean-predfsd,ymax=predmean+predfsd), alpha=0.4) +
  geom_point(aes(y=obs)) +
  theme_bw()
```

### Plot conditional responses

The function `getconditionals` will compute and plot conditional responses to each input variable (other input varibles set to their mean value). From this we can also clearly see that lags 2 and 3 have no impact, and we can see how the lag 1 dynamics of the 2 populations differ.

```{r conditionals}
con=getconditionals(tlogtest)
```

If you prefer ggplot:

```{r ggplot-conditionals}
#have to convert conditionals output to long format
#there may be a more concise way to do this
library(tidyr)
npreds=length(grep("_yMean",colnames(con)))
conlong1=gather(con[,1:(npreds+1)],x,xValue,2:(npreds+1))
conlong2=gather(con[,c(1,(npreds+2):(2*npreds+1))],ym,yMean,2:(npreds+1))
conlong3=gather(con[,c(1,(2*npreds+2):(3*npreds+1))],ys,ySD,2:(npreds+1))
conlong=cbind.data.frame(conlong1,yMean=conlong2$yMean,ySD=conlong3$ySD)
ggplot(conlong,aes(x=xValue,y=yMean)) +
  facet_grid(pop~x, scales = "free") +
  geom_line() + geom_ribbon(aes(ymin=yMean-ySD,ymax=yMean+ySD), alpha=0.4) +
  theme_bw()
```

### Plot inverse length scales

The model hyperparameters are located under `model$pars`. If you have $n$ predictors, the first $n$ values of `pars` will be the length scales. Note that if you use `E` and `tau`, the names of the predictors in the input data frame will be stored under `model$inputs$x_names`, and the names of the lagged predictors corresponding to the inverse length scales will be stored under `model$inputs$x_names2`.

```{r ls, fig.height=3}
predvars=tlogtest$inputs$x_names2
npreds=length(predvars)
lscales=tlogtest$pars[1:npreds]
par(mar=c(4,4,1,1))
plot(factor(predvars),lscales,xlab="Predictor",ylab="Inverse length scale")
```

### Other types of predictions

We can use the `predict` function to get various types of predictions from a fitted model. Leave-one-out, `predictmethod = "loo"` is one option (above, we got these predictions at the same time we fit the model, but this is not necessary). The following obtains sequential (leave-future-out) predictions using the training data. Note that for `predictmethod = "loo"` and `predictmethod = "sequential"`, the training data are iteratively omitted for the predictions, but the hyperparameter values used are those obtained using *all* of the training data (the model is not refit).

```{r predict}
#sequential predictions (they should improve over time)
seqpred=predict(tlogtest,predictmethod = "sequential")
plot(seqpred)
```

You could, alternatively, supply new data for which to make predictions. In that case, you would supply a new data frame (`newdata`). In our example, this data frame which should contain columns `Abundance` and `Population`.  

A common approach when fitting these models is to split the available data into a training and test dataset. For instance, say we wanted a single-population model for PopA with 2 time lags, and we wanted to use the first 40 data points as training data, and the last 10 points as test data. For that we could do the following.

```{r predict2, fig.height=3}
pAtrain=pA[1:40,]
pAtest=pA[41:50,]
tlogtest2=fitGP(data = pAtrain, y = "Abundance", E=2, tau=1, time = "Time", newdata = pAtest)
plot(tlogtest2)
```

If you don't want the first `E*tau` points getting excluded from your test data, generate the lags beforehand, then split the data (don't use `E` and `tau` options in `fitGP`). See [Specifying training data] (option 1).

```{r predict3, fig.height=3}
pAlags=makelags(pA, y = "Abundance", E=2, tau=1)
pAdata=cbind(pA,pAlags)
pAtrain=pAdata[1:40,]
pAtest=pAdata[41:50,]
tlogtest3=fitGP(data = pAtrain, y = "Abundance", x=colnames(pAlags), time = "Time", newdata = pAtest)
plot(tlogtest3)
```

Note that this is equivalent to:

```{r predict3b}
tlogtest3=fitGP(data = pAtrain, y = "Abundance", x=colnames(pAlags), time = "Time")
tlogtest3_pred=predict(tlogtest3, newdata = pAtest)
```

### Simulating forecasts

The function `predict_seq` also generates predictions for a data frame `newdata`, but sequentially adds each new observation to the training data and refits the model. This simulates a real-time forecasting application. This prediction method is similar to `predictmethod="sequential"` (leave-future-out), but in this case, the future time points are not included in the training data used to obtain the hyperparameters. It is also similar to the train/test split using `newdata`, but in this case, the training data and model hyperparameters are sequentially updated with each timestep. Use of this method requires the use of `data` with pre-generated lags (see [Specifying training data], option A1), that a `time` column is specified, that `newdata` has exactly the same columns as `data`, and that `newdata` has observed values.  

```{r predict4, fig.height=3}
tlogtest3_update=predict_seq(tlogtest3, newdata=pAtest)

summary(tlogtest3)

summary(tlogtest3_update)

plot(tlogtest3_update)
```

### Making forecasts

You can also use `predict` to generate predictions beyond the end of the training data. You can construct a forecast matrix extending beyond the last timepoint using `makelags` and setting `forecast=T`. This can be supplied this as `newdata`. To use this method, you have to generate the lags beforehand for *both* the training data and the forecast (use `makelags` for both, and the settings in `makelags` should match, other than `forecast`, see [Specifying training data] option 1). It is a good idea to include the `time` argument when doing this.

```{r forecast, fig.height=3}
lags1=makelags(thetalog2pop,y=c("Abundance"),pop="Population",time="Time",E=3,tau=1)
fore1=makelags(thetalog2pop,y=c("Abundance"),pop="Population",time="Time",E=3,tau=1,forecast = T)
data1=cbind(thetalog2pop, lags1)

tail(data1)

fore1

tlogfore=fitGP(data = data1, y = "Abundance", x=c("Abundance_1","Abundance_2","Abundance_3"), 
               pop = "Population", time = "Time", scaling = "local", newdata = fore1)

ggplot(tlogfore$insampresults,aes(x=timestep,y=predmean)) +
  facet_wrap(pop~., scales = "free") +
  geom_line() + geom_ribbon(aes(ymin=predmean-predsd,ymax=predmean+predsd), alpha=0.4) +
  geom_point(aes(y=obs)) +
  geom_point(data=tlogfore$outsampresults, aes(y=predmean), color="red") +
  geom_errorbar(data=tlogfore$outsampresults,
                aes(ymin=predmean-predsd,ymax=predmean+predsd),color="red") +
  theme_bw()
```

### Making iterated forecasts

The `predict_iter` function can be used to obtain iterated predictions from `fitGP` models with lag predictors. This method also requires pre-generated lags and the use of `newdata` (see [Specifying training data] option 1). The `time` argument is required.

Starting with the first row of `newdata`, the predicted `y` variable is inserted as the first lag for the next timestep, the other lags are shifted right by 1, and a new prediction is made. This continues for as many rows as are in `newdata`. Thus, this procedure assumes that all lags are present, evenly spaced, and in order (first lag is the first column). Time steps are assumed to be evenly spaced. The units of `y` and its lags are assumed to be the same.

Unless otherwise specified, all values of `x` are assumed to be time lags of `y`. If this is not the case (the model contains a mixture of lags and covariates), you must specify which columns are the time lags (argument `xlags` in `predict_iter`). The exception is for "fisheries models" (`fitGP_fish`), where the lags and covariates are already separately specified.

As with other prediction methods, `newdata` should contain all values of `x`, as well as `time`. Only the first row of lags in `newdata` (for each population) needs to be completely filled in. Subsequent time points can be NA or filled in - they will be overwritten with the predictions as the model iterates. The values of `time` and must be filled in. If the model has covariates, values for the covariates need to be supplied for all timepoints in `newdata`. If there are multiple populations, `newdata` should contain `pop` and there should be duplicated timesteps for each population.

As expected for a chaotic system, the iterated predictions from the simulated 2 population theta logistic model become very poor after a few timesteps.

```{r itergeneral}
xdata=makelags(data = thetalog2pop, y="Abundance", pop="Population", time = "Time", E=3, tau=1, append=T)
xdatatrain=subset(xdata, Time<=40)
#get iterated predictions for the last 10 timesteps
xdatafore=subset(xdata, Time>40)

tlog=fitGP(data = xdatatrain, y = "Abundance", x=c("Abundance_1","Abundance_2","Abundance_3"), 
           pop = "Population", time = "Time", scaling = "local")
predtest=predict_iter(tlog, newdata = xdatafore)

plot(predtest)
```

If you want to iterate forecasts beyond the end of the time series, here is how you might set that up. Note that because there are no observed values, we can't evaluate fit or make plots of observed vs. predicted.

```{r itergeneral2}
xdata=makelags(data = thetalog2pop, y="Abundance", pop="Population", time="Time", E=3, tau=1, append=T)
#use forecast feature to get first rows
xdatafore=makelags(data = thetalog2pop, y="Abundance", pop="Population", time="Time", E=3, tau=1, forecast=T)
#create remaining timepoints for each population
xdatafore2=expand.grid(Time=52:65, Population=unique(thetalog2pop$Population))
#create empty matrix for future values
xdatafore3=matrix(NA, nrow=nrow(xdatafore2), ncol=3,
                  dimnames=list(NULL,c("Abundance_1","Abundance_2","Abundance_3")))
#combine everything
xdatafore=rbind(xdatafore,cbind(xdatafore2,xdatafore3))

head(xdatafore)

tlog2=fitGP(data = xdata, y = "Abundance", x=c("Abundance_1","Abundance_2","Abundance_3"), 
           pop = "Population", time = "Time", scaling = "local")
predtest2=predict_iter(tlog2, newdata = xdatafore)
#plot(predtest2)
head(predtest2$outsampresults)
```


## Specifying training data

There are several ways that the training data for a model can be specified. 

A. supply data frame `data`, plus column names or indices for `y` and `x`.  
B. supply a vector for `y` and a vector or matrix for `x`.  

For each of the above 2 options, there are 3 options for specifying the predictors.  

1. supplying `y` and `x` (omitting `E` and `tau`) will use the columns of `x` as predictors. This allows for the most customization.  
2. supplying `y`, `E`, and `tau` (omitting `x`) will use `E` lags of `y` (with spacing `tau`) as predictors. This is equivalent to option 3 with `x`=`y`.  
3. supplying `y`, `x`, `E`, and `tau` will use `E` lags of *each column* of `x` (with spacing `tau`) as predictors. Do not use this option if `x` already contains lags, in that case use option 1.  

The above examples use methods A2 and A1. 

Option 1 allows for the most customization of response and predictor variables, including use of mixed embeddings, and use of different variables for the response and predictors. Options 2 and 3 exist for convenience, but for the most control over the model and to use the more elaborate features in this package, it is best to use option 1: use `makelags` to generate any lags beforehand and pass appropriate columns to `fitGP`, rather than rely on `fitGP` to generate lags internally. Option A will make more sense if your data are already in a data frame, option B may make more sense if you are doing simulations and just have a bunch of vectors and matrices.

The `pop` argument is optional in all of the above cases. Beware that if omitted, a single population is assumed.

```{r other_methods}
set.seed(10)
thetalog2pop$othervar=rnorm(nrow(thetalog2pop))
yvec=thetalog2pop$Abundance
popvec=thetalog2pop$Population
#function 'makelags' can be used to generate a lag matrix
#be sure to include 'pop' if data contain multiple pops to prevent crossover
xmat=makelags(y=thetalog2pop[,c("Abundance","othervar")],pop=popvec,E=2,tau=1)
thetalog2pop2=cbind(thetalog2pop,xmat)

#Method A1
ma1=fitGP(data=thetalog2pop2,y="Abundance",x=c("Abundance_1","othervar"),
         pop="Population",scaling="local")

#Method B1
#like A1, but your data aren't in a data frame
mb1=fitGP(y=yvec,x=xmat,pop=popvec,scaling="local")

#Method B2
#like A2, but your data aren't in a data frame
mb2=fitGP(y=yvec,pop=popvec,E=2,tau=1,scaling="local")

#Method A3
#generate lags of multiple predictors internally
ma3=fitGP(data=thetalog2pop2,y="Abundance",x=c("Abundance","othervar"),
         pop="Population",E=2,tau=1,scaling="local")


summary(ma1)

summary(mb1)

summary(mb2)

summary(ma3)
```

## Extensions

See [this vignette](https://tanyalrogers.github.io/GPEDM/articles/vtimestep.html) for more detail about the variable timestep method (VS-EDM) for missing data. 

See [this vignette](https://tanyalrogers.github.io/GPEDM/articles/fisheries.html) for more detail on an alternative parameterization of the GP model, `fitGP_fish`, designed for use in fisheries applications.

## References

Munch, S. B., Poynor, V., and Arriaza, J. L. 2017. Circumventing structural uncertainty: a Bayesian perspective on nonlinear forecasting for ecology. Ecological Complexity, 32:134.

Johnson, B., and Munch, S. B. 2022. An empirical dynamic modeling framework for missing or irregular samples. Ecological Modelling, 468:109948.


*Any advice on improving this package is appreciated.*

****

```{r add-disclaimer, echo=FALSE, results='asis'}
url <- "https://raw.githubusercontent.com/nmfs-fish-tools/Resources/master/Disclaimer.md"
childtext <- readLines(url,encoding = "UTF-8")
cat(childtext, sep="\n")
```
