---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# GPEDM

<!-- badges: start -->
<!-- badges: end -->

**Disclaimer: This is a very much a work in progress. Use at your own risk.**

This package contains functions for fiting hierarchical, separable length scale GP models with automatic relevance determination (ARD) for use in Empirical Dynamic Modeling (EDM) and other applications. This is an adaptation of code originally developed by Stephan Munch in MATLAB.  

The main function is `fitGP` which is used to train the model and can also produce predictions if desired. Use `predict.GP` to generate other or additional predictions from a fitted model, `plot.GP` to plot observed and predicted values and `getconditionals` to obtain conditional reponses. Also available for use is the function `makelags` which can be used to create delay vectors. See the (not yet created) vignette for more detailed instructions.

## Installation

To install the package:

``` r
devtools::install_github("tanyalrogers/GPEDM")
```
If you are using Windows, you may need to install [Rtools](https://cran.r-project.org/bin/windows/Rtools/).

## Simple Example

These are truly trivial examples that I should replace with something else. I was using them mainly for testing.

```{r example}
library(GPEDM)
set.seed(10)

#some trival data
l=20
x1=seq(-5,5,length.out = l)
x2=rnorm(l,0,1)
y=0.5*x1^2+rnorm(l)
xmat=cbind(x1,x2)
#create a missing value
ym=y
ym[5]=NA
#as a data frame
dftest=cbind.data.frame(response=y,responsem=ym,xmat,pop=rep(c(1,2),10),pop2=rep(c("site A","site B"),10))
dftest2=dftest[order(dftest$pop),]

#specify training data as vector and matrix
test=fitGP(yd=y,xd=xmat,scaling = "global",predictmethod = "loo")
#see a summary
summary(test)
#plot out of sample results
plot(test)
#plot in sample results
plot(test, plotinsamp = T)
#plot conditional responses
con1=getconditionals(test, plot = T)

#specify training data with data frame, example with 2 populations and a missing data point
testhier=fitGP(data=dftest2,yd="responsem",xd=c("x1","x2"),pop="pop2",scaling="local", predictmethod="loo")
summary(testhier)
plot(testhier)
cond2=getconditionals(testhier)

#specify training data as lags (E, tau)
yrand = rnorm(50,0,1)
testrand=fitGP(yd=yrand,E=2,tau=1)
#alternatively
yrandlags=makelags(yrand,E=2,tau=1)
testrand2=fitGP(yd=yrand,xd=yrandlags)

```

## References

Munch, S. B., Poynor, V., and Arriaza, J. L. 2017. Circumventing structural uncertainty: a Bayesian perspective on nonlinear forecasting for ecology. Ecological Complexity, 32:134.


*Any advice on improving this package is appreciated.*
