---
title: "GPEDM Models For Fisheries"
output: rmarkdown::html_vignette
toc: true
vignette: >
  %\VignetteIndexEntry{GPEDM Models For Fisheries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(GPEDM)
library(ggplot2)
```

## Introduction

This vignette discusses the use of an alternative parameterization of the GP-EDM model designed for use in fisheries applications, implemented in the function `fitGP_fish`. This fits a GP model of the form
$$y=f(m-bh,z)$$
where $y$ is an index of abundance (typically in units of catch per unit effort, CPUE), $m$ are lags of $y$, $h$ are lags of harvest (in biomass or numbers), and $z$ are optional covariates. The index of abundance is assumed proportional to biomass, with proportionality constant $b$ (the "catchability", units: CPUE/biomass). The composite variable $m-bh$ is assumed proportional to the biomass of individuals remaining after harvesting ("escapement").

The function `fitGP_fish` finds parameter $b$ using `stats::optimize` applied to the posterior likelihood of fitted GP models given $b$. If you want to skip optimization of $b$ and use a fixed value for it, one can be provided under `bfixed`.

The function `fitGP_fish` has all of the same functionality as `fitGP`, including the use of hierarchical structures and augmentation data, and can be used with all of the `predict` functions in the package. In all prediction cases, it is only necessary to supply $m$ and $h$ - escapement will be calculated internally. The inputs and structure of the inputs are somewhat more constrained than in `fitGP` (see below), which is just something to be aware of.

## Sample dataset

As a sample dataset, we will use a simulated Ricker model with harvest. This model has chaotic dynamics, and data from two 'regions'. For this first example we will just use Region A, where we know the true value of $b$ is 0.01. 

```{r sampledata, fig.height=3}
data("RickerHarvest")
RickerHarvestA=subset(RickerHarvest, Region=="A")
RickerHarvestB=subset(RickerHarvest, Region=="B")
par(mfrow=c(1,2),mar=c(4,4,2,1))
plot(Catch~Time, data=RickerHarvestA, type="l", main="Region A")
plot(CPUE_index~Time, data=RickerHarvestA, type="l", main="Region A")
```

## Fitting a model

The function `fitGP_fish` requires the use of `data` with pre-generated lags (option A1 in [Specifying training data](https://tanyalrogers.github.io/GPEDM/index.html#specifying-training-data))). Values for `y`, `m`, and `h` are required, which should be the names of the appropriate columns in `data`. A value for `time` is also recommended. For the sample dataset, we will use just one lag.

```{r modelfit1}
RHlags=makelags(RickerHarvestA, y=c("CPUE_index","Catch"), time="Time", tau=1, E=1, append=T)

fishfit0=fitGP_fish(data=RHlags, y="CPUE_index", m="CPUE_index_1", h="Catch_1", time="Time")
summary(fishfit0)
```

Computed values of "escapement" (lagged) will be appended to the `model$insampresults` table (also to any `model$outsampresults` table). By default, the composite variable will be called "escapement", unless you supply a different name for it (argument `xname` of `fitGP_fish`). In this case, no data transformation is applied (see [Data tranformations](#data-transformations), so `predmean_trans` and `predmean` are the same.

```{r}
head(fishfit0$insampresults)
```

Conditional effects (from `getconditionals`) will be plotted with respect to the calculated escapement values (and any covariates, if present).

```{r conditionals1}
getconditionals(fishfit0)
```

Note that this function does not go through (0,0) because there are no data there. This could potentially cause problems if the model is extrapolated to regions of low or no escapement (as a result of high catch rates).

To force the model through the origin (so that zero escapement means zero CPUE), you can supply an additional datapoint at 0 escapement, 0 CPUE. It works well to supply these under `augdata`, since these points are used as training data and in determining the range the conditional plots, but will not be included in the results table, used to evaluate fit, or plotted as part of the time series. Note that the model will probably not go through 0 exactly, because error is assumed the same at the origin as it is elsewhere. Also note that if a population experiences substantial immigration or emigration, the true function may not actually intersect the origin.

```{r modelfit2}
#force zero escapement = zero CPUE
#the value for Time doesn't matter, but it should NOT be an NA
zeropin=data.frame(Time=0,CPUE_index=0,CPUE_index_1=0,Catch_1=0)

fishfit=fitGP_fish(data=RHlags, y="CPUE_index", m="CPUE_index_1", h="Catch_1", time="Time", 
                   augdata=zeropin)
summary(fishfit)

getconditionals(fishfit)
```

## Multiple populations

If you have multiple regions (populations), each with their own CPUE index and Catch values, you can fit a hierarchical model using `fitGP_fish` in the same way as `fitGP`. The default behavior is to assume that $b$ is the same for both population (`bshared=TRUE`). Supplying a single value under `bfixed` will use the same fixed value for all populations. 

To estimate separate $b$ values for each population, set `bshared=FALSE`. This will use the Nelder-Mead method of `stats::optim`, which will be quite a bit slower than the single $b$ optimization, so more than 3 populations would not be recommended. You can fix different values of $b$ for each population by supplying a named vector under `bfixed`.

If using different values of $b$, be sure to use `scaling="local"` because the values are assumed to be in different units by virtue of having different catchabilities, but the dynamics should be proportional. You may want to check whether the combined model actually produces better within-population fits than the models fit separately. If there are sufficient data, there is a good chance it will not.

The sample dataset contains time series for a second region B, where the true value of $b$ is 0.005. It also has a different catch history.

```{r sampledataB, fig.height=3}
par(mfrow=c(1,2),mar=c(4,4,2,1))
plot(Catch~Time, data=RickerHarvestB, type="l", main="Region B")
plot(CPUE_index~Time, data=RickerHarvestB, type="l", main="Region B")
```

A separate model fith to region B:

```{r modelfitB}
#model for region B
RHlagsB=makelags(RickerHarvestB, y=c("CPUE_index","Catch"), time="Time", tau=1, E=1, append=T)
zeropin=data.frame(Time=0,CPUE_index=0,CPUE_index_1=0,Catch_1=0)

fishfitB=fitGP_fish(data=RHlagsB, y="CPUE_index", m="CPUE_index_1", h="Catch_1", time="Time",
                    augdata=zeropin)
summary(fishfitB)
```

A hierarchical model assuming a single $b$ value:

```{r hier1}
RHlags2pop=makelags(RickerHarvest, y=c("CPUE_index","Catch"), time="Time", 
                    pop = "Region", tau=1, E=1, append=T)
zeropin=data.frame(Region=c("A","B"),Time=0,CPUE_index=0,CPUE_index_1=0,Catch_1=0)

#model assuming a shared b
fishfit2pop1=fitGP_fish(data=RHlags2pop, y="CPUE_index", m="CPUE_index_1", h="Catch_1",
                        pop="Region", time="Time", scaling = "local", augdata = zeropin)
summary(fishfit2pop1)
```

A hierarchical model assuming separate values of $b$ (estimated):

```{r hier2}
#model assuming separate b values (takes a while to run)
fishfit2pop2=fitGP_fish(data=RHlags2pop, y="CPUE_index", m="CPUE_index_1", h="Catch_1",
                        pop="Region", time="Time", scaling = "local", augdata = zeropin,
                        bshared = FALSE)
summary(fishfit2pop2)
```

A hierarchical model assuming separate values of $b$ (fixed):

```{r hier3}
#model assuming separate b values (fixed values)
b=c(A=0.01, B=0.005)
fishfit2pop2f=fitGP_fish(data=RHlags2pop, y="CPUE_index", m="CPUE_index_1", h="Catch_1",
                        pop="Region", time="Time", scaling = "local", augdata = zeropin,
                        bfixed = b)
summary(fishfit2pop2f)
getconditionals(fishfit2pop2f)
```

If more flexibility is required or you want to do something more elaborate, below is code for how you could calculate escapement externally given some value(s) of $b$, and fit a regular `fitGP` to the escapement values. The example below fits the same model as above (with $b$ fixed).

```{r combomodel, fig.height=3}
bA=0.01
bB=0.005
RHlags$escapement_1=RHlags$CPUE_index_1-RHlags$Catch_1*bA
RHlagsB$escapement_1=RHlagsB$CPUE_index_1-RHlagsB$Catch_1*bB

par(mfrow=c(1,2),mar=c(4,4,2,1))
plot(CPUE_index~escapement_1, data=RHlags, main="Region A")
plot(CPUE_index~escapement_1, data=RHlagsB, main="Region B")
```

```{r combomodel2}
RHlagscombo=rbind(RHlags, RHlagsB)
zeropin=data.frame(Region=c("A","B"),Time=0,CPUE_index=0,escapement_1=0)

fishfitcombo=fitGP(data=RHlagscombo, y="CPUE_index", x="escapement_1", time="Time",
                   pop="Region", scaling="local", augdata=zeropin)
summary(fishfitcombo)
```

## Getting MSY using iterated prediction

Given a fitted fisheries model, the steady state catch and biomass given a certain harvest rate can be obtained using iterated prediction. This can be done using the `predict_iter` function. Similar to `predict`, you supply a `newdata` data frame, which (for a one-population model) should contain the `time`, `m`, and `h` columns. You also supply a harvest rate `hrate` from 0 to 1, which indicates the proportion of predicted CPUE biomass to be harvested. 

Starting with the first row of `newdata`, the predicted `y` variable is inserted into the first column of `m` for the next timestep, and the other values of `m` are shifted right by 1. The first value of `h` at the next timestep is calculated as `y/b*hrate` and the other values of `h` are shifted right by 1. Escapement is recalculated, and a new prediction for `y` is made. This continues for as many rows as are in `newdata`. Thus, this procedure assumes that all lags are present, evenly spaced, and in order (first lag is the first column). Time steps are assumed to be evenly spaced. The units of `y` and `m` are assumed to be the same.

In `newdata`, only the first timepoint (row) for `m` and `h` needs to be filled in. The rest can be NA and will be filled in as the model iterates forward. The values of `time` must be filled in. If there are covariates (`z`), they should also be included in `newdata`, and values must be supplied for all time points. If there are multiple populations, `newdata` should contain `pop` and there should be duplicated timesteps for each population (see the next sections for a 2-population example).

```{r msy1, fig.height=3}
#number of timesteps to iterate
nfore=50

#use the forecast feature to create the first row
RHfore1=makelags(RickerHarvestA, y=c("CPUE_index","Catch"), time="Time", tau=1, E=1, forecast=T)
#get remaining timepoints. use expand.grid here if you have multiple populations.
RHfore2=data.frame(Time=(RHfore1$Time[1]+1):(RHfore1$Time[1]+nfore-1)) 
#create empty matrix for future values
RHfore3=matrix(NA, nrow=nrow(RHfore2), ncol=2,
               dimnames=list(NULL,c("CPUE_index_1","Catch_1")))
#combine everything
RHfore=rbind(RHfore1,cbind(RHfore2,RHfore3))

head(RHfore)
```

Since that is kind of annoying, I have written a wrapper for it called `makelags_iter`. The arguments are the same as `makelags` but you specify `nfore` and it makes the matrix above.

```{r}
RHfore=makelags_iter(nfore=50, data=RickerHarvestA, y=c("CPUE_index","Catch"), 
                     time="Time", tau=1, E=1)
head(RHfore)
```

This matrix can be passed to `predict_iter`, specifying a harvest rate.

```{r}
msyfore=predict_iter(fishfit, newdata = RHfore, hrate = 0.5)
```

The output of `predict_iter` can be passed to `plot` if desired, which will plot CPUE (y). Note that there will not be any observed values. The resulting escapement and catch values will be included in the `outsampresults` table.

```{r msy2, fig.height=3}
plot(msyfore)

head(msyfore$outsampresults)
```

For a given harvest rate, may be valuable to plot the projection alongside the past values.

```{r msyplots1, fig.height=3}
#CPUE
ggplot(msyfore$outsampresults, aes(x=timestep, y=predmean)) +
  geom_line(size=1) + #iterated predictions
  geom_line(data=fishfit$insampresults) + ylab("CPUE") + #past predictions
  geom_point(data=fishfit$insampresults, aes(y=obs)) #observed values
#Catch
ggplot(msyfore$outsampresults, aes(x=timestep, y=Catch_1)) +
  geom_line(size=1) +
  geom_line(data=RHlags, aes(x=Time)) #past catches
#Escapement
ggplot(msyfore$outsampresults, aes(x=timestep, y=escapement_1)) +
  geom_line(size=1) +
  geom_line(data=fishfit$insampresults) #past escapements
```

Evaluating the steady state catch across a range of `hrate` values allows you to obtain the maximum sustainable yield (MSY), associated fishing rate (f_MSY), and associated CPUE indexed biomass (proportional to B_MSY). In the example below, we make iterated predictions for 50 timesteps as above, but record just the last 10 values. Since the resulting time series in this example can exhibit chaos and cycles, we then take the average over those values for each harvest rate.

```{r msy3, fig.height=3}
#vector of harvest rates
hratevec=seq(0,1,length.out=30)
#number of timepoints to save
tsave=10

catchsave=expand.grid(time=1:tsave,hrate=hratevec)
catchsave$catch=NA
catchsave$cpue=NA
for(i in seq_along(hratevec)) {
  msyfore=predict_iter(fishfit, newdata = RHfore, hrate = hratevec[i])
  #note that the next 2 lines only work if there is one population
  catchsave$catch[catchsave$hrate==hratevec[i]]=
    msyfore$outsampresults$Catch_1[(nfore-tsave+1):nfore]
  catchsave$cpue[catchsave$hrate==hratevec[i]]=
    msyfore$outsampresults$predmean[(nfore-tsave+1):nfore]
}

#average over the last 10 timepoints
catchsavemean=aggregate(cbind(catch,cpue)~hrate, data=catchsave, mean)
(fmsy=catchsavemean$hrate[which.max(catchsavemean$catch)])
(Bmsy=catchsavemean$cpue[which.max(catchsavemean$catch)])

par(mfrow=c(1,2),mar=c(4,4,2,1))
plot(catch~hrate, data=catchsave, col="gray")
lines(catch~hrate, data=catchsavemean, lwd=2, col="red")
abline(v=fmsy, col="red")
plot(cpue~hrate, data=catchsave, col="gray")
lines(cpue~hrate, data=catchsavemean, lwd=2, col="red")
abline(v=fmsy, col="red")
```

Here's a wrapper for all of that as well. Since there is only one population here, the 'total' and 'pop' output tables are identical. The ones you really care about are `catchforeall` which are all the predictions, `catchsavetotal` which are the last `tsave` predictions, and `catchsavetotalmean` which is the average of the last `tsave` predictions.

```{r msy3.5, fig.height=3}
hratevec=seq(0,1,length.out=30)
msyout=msy_wrapper(model=fishfit, newdata = RHfore, hratevec = hratevec, tsave = 10)
msyout$fmsy
msyout$Bmsy

par(mfrow=c(1,2),mar=c(4,4,2,1))
plot(catch~hrate, data=msyout$catchsavetotal, col="gray")
lines(catch~hrate, data=msyout$catchsavetotalmean, lwd=2, col="red")
abline(v=msyout$fmsy, col="red")
plot(cpue~hrate, data=msyout$catchsavetotal, col="gray")
lines(cpue~hrate, data=msyout$catchsavetotalmean, lwd=2, col="red")
abline(v=msyout$fmsy, col="red")
```

Let's look at the trajectory for the population at fmsy.

```{r msyplots2, fig.height=3}
#you can get this either by rerunning predict_iter
msytraj=predict_iter(fishfit, newdata = RHfore, hrate = fmsy)$outsampresults
#or if you have used the msy_wrapper, you can get it out or the catchforeall table
msytraj=subset(msyout$catchforeall, hrate==fmsy)

#CPUE
ggplot(msytraj, aes(x=timestep, y=predmean)) +
  geom_line(size=1) + #iterated predictions
  geom_line(data=fishfit$insampresults) + ylab("CPUE") + #past predictions
  geom_point(data=fishfit$insampresults, aes(y=obs)) #observed values
#Catch
ggplot(msytraj, aes(x=timestep, y=Catch_1)) +
  geom_line(size=1) +
  geom_line(data=RHlags, aes(x=Time))
#Escapement
ggplot(msytraj, aes(x=timestep, y=escapement_1)) +
  geom_line(size=1) +
  geom_line(data=fishfit$insampresults)
```

Let's compare this to the true analytical solution for this model using the simulation parameters.

```{r msy4, fig.height=3}
r=3; K=1000
q=0.01 #catchability b
tbiomass=K/(1-hratevec)*(r-log(1/(1-hratevec))) #steady state biomass
tcpue=q*tbiomass
tcatch=hratevec*tbiomass

par(mfrow=c(1,2),mar=c(4,4,2,1))
plot(catch~hrate, data=msyout$catchsavetotalmean, type="l", lwd=2, col="red")
lines(tcatch~hratevec, lwd=2, col="cornflowerblue")
abline(v=msyout$fmsy, col="red")
legend(x="topleft", legend = c("EDM","true"), col=c("red","cornflowerblue"), lwd=2, cex=0.8)
plot(cpue~hrate, data=msyout$catchsavetotalmean, type="l", lwd=2, col="red")
lines(tcpue~hratevec, lwd=2, col="cornflowerblue")
abline(v=msyout$fmsy, col="red")
legend(x="topleft", legend = c("EDM","true"), col=c("red","cornflowerblue"), lwd=2, cex=0.8)
```

It is also possible to set this up using a constant catch value rather than constant catch rate. To do this, omit `hrate` and fill all the catch rows and columns in `newdata` with a constant value. In this example with chaotic dynamics, this can lead to negative escapement, so it was not done here.

## MSY with multiple populations

If you have a multi-population `fitGP_fish` model that assumes a single values of $b$, or if you are using separate `fitGP_fish` models for each population, you should be able to use `predict_iter` as above, but will need to keep track of the predictions for each population and decide if you want to add them together or not.

At present, the harvest rate is assumed the same for all populations. This may change in the future.

```{r msy2pop1}
#make the forecast matrix
RHfore2pop=makelags_iter(nfore=50, data=RickerHarvest, y=c("CPUE_index","Catch"), 
                         time="Time", tau=1, E=1, pop="Region")
head(RHfore2pop)
```

Plots of the trajectories for a given harvest rate:

```{r msyplots3, fig.height=3}
msyfore2pop=predict_iter(fishfit2pop2, newdata = RHfore2pop, hrate = 0.5)
head(msyfore2pop$outsampresults)

#CPUE
ggplot(msyfore2pop$outsampresults, aes(x=timestep, y=predmean, color=pop, group=pop)) +
  geom_line(size=1) +
  geom_line(data=fishfit2pop2$insampresults) + ylab("CPUE") +
  geom_point(data=fishfit2pop2$insampresults, aes(y=obs))
#Catch
ggplot(msyfore2pop$outsampresults, aes(x=timestep, y=Catch_1, color=pop, group=pop)) +
  geom_line(size=1) +
  geom_line(data=RHlags2pop, aes(x=Time, color=Region, group=Region))
#Escapement
ggplot(msyfore2pop$outsampresults, aes(x=timestep, y=escapement_1, color=pop, group=pop)) +
  geom_line(size=1) +
  geom_line(data=fishfit2pop2$insampresults)
```

Evaluate a grid of harvest rates to obtain MSY using the wrapper. Since there are multiple populations, the 'total' and 'pop' tables will be different. The MSY estimates reported by the function are based on `catchsavetotalmean`, but you could compute population specific estimates if desired.

```{r msy2pop2, fig.height=3}
hratevec=seq(0,1,length.out=30)
msyout2pop=msy_wrapper(model=fishfit2pop2, newdata = RHfore2pop, hratevec = hratevec, tsave = 10)
msyout2pop$fmsy
msyout2pop$Bmsy

#Total
par(mfrow=c(1,2),mar=c(4,4,2,1))
plot(catch~hrate, data=msyout2pop$catchsavetotal, col="gray")
lines(catch~hrate, data=msyout2pop$catchsavetotalmean, lwd=2, col="red")
abline(v=msyout2pop$fmsy, col="red")
plot(cpue~hrate, data=msyout2pop$catchsavetotal, col="gray")
lines(cpue~hrate, data=msyout2pop$catchsavetotalmean, lwd=2, col="red")
abline(v=msyout2pop$fmsy, col="red")

#Split up by pop
ggplot(msyout2pop$catchsavepop, aes(x=hrate, y=catch, color=pop)) +
  geom_point(alpha=0.5) +
  geom_line(data=msyout2pop$catchsavepopmean, size=1) +
  geom_vline(xintercept = msyout2pop$fmsy, color="black")
ggplot(msyout2pop$catchsavepop, aes(x=hrate, y=cpue, color=pop)) +
  geom_point(alpha=0.5) +
  geom_line(data=msyout2pop$catchsavepopmean, size=1) +
  geom_vline(xintercept = msyout2pop$fmsy, color="black")
```

If you are computing escapement externally, you will probably have to write a custom loop, like below. This is mostly the internal code of `predict_iter` but with some modification.

```{r combomsy}
#calculate escapement
RHfore2pop$b=ifelse(RHfore2pop$Region=="A", bA, bB)
RHfore2pop$escapement_1=RHfore2pop$CPUE_index_1-RHfore2pop$Catch_1*RHfore2pop$b

head(RHfore2pop)

#vector of harvest rates
hratevec=seq(0,1,length.out=40)
#number of timepoints to save
tsave=10

#rename some stuff to minimize recoding
popname="Region"
timename="Time"
object=fishfitcombo
xlags="CPUE_index_1"
hlags="Catch_1"
newtimes=unique(RHfore2pop[,timename])
up=unique(RHfore2pop[,popname])

#setup final data frame
catchsavepop=expand.grid(pop=up,time=1:tsave,hrate=hratevec)
catchsavepop$catch=NA
catchsavepop$cpue=NA

pred=list()

for(h in seq_along(hratevec)) {
  
  newdata=RHfore2pop
  hrate = hratevec[h]
  
  #iterated prediction loop
  for(i in seq_along(newtimes)) {
    newdatai=newdata[newdata[,timename]==newtimes[i],]
    #get prediction
    pred[[i]]=predict(object, newdata=newdatai)$outsampresults
    preds=pred[[i]]
    #create new data row for each population
    #assumes all lags are present and evenly spaced
    if(i+1 <= length(newtimes)) {
      for(j in 1:length(up)){
        predsj=preds[preds$pop==up[j],]$predmean
        thisrowj=newdata[,timename]==newtimes[i] & newdata[,popname]==up[j]
        nextrowj=newdata[,timename]==newtimes[i+1] & newdata[,popname]==up[j]
        newdata[nextrowj,xlags[1]]=predsj
        if(length(xlags)>1) {
          newdata[nextrowj,xlags[2:length(xlags)]]=newdata[thisrowj,xlags[1:(length(xlags)-1)]]
        }
        #compute next catch
        newdata[nextrowj,hlags[1]]=predsj/newdata[nextrowj,"b"]*hrate
        if(length(xlags)>1) {
          newdata[nextrowj,hlags[2:length(xlags)]]=newdata[thisrowj,hlags[1:(length(xlags)-1)]]
        }
      }
      #compute escapement
      newdata$escapement_1=newdata[,xlags]-newdata[,hlags]*newdata[,"b"]
    }
  }
  
  #combine results
  msyfore=do.call(rbind, pred)
  msyfore=cbind(msyfore,newdata[,hlags,drop=F])
  
  #pull out last tsave timepoints for each population
  for(p in 1:length(up)){
    msyforei=subset(msyfore, pop==up[p])
    nfore=nrow(msyforei)
    catchsavepop$catch[catchsavepop$hrate==hratevec[h] & catchsavepop$pop==up[p]]=
      msyforei[(nfore-tsave+1):nfore,hlags[1]]
    catchsavepop$cpue[catchsavepop$hrate==hratevec[h] & catchsavepop$pop==up[p]]=
      msyforei$predmean[(nfore-tsave+1):nfore]
  }
}

#sum over pops (if more than one)
catchsavetotal=aggregate(cbind(catch,cpue)~hrate*time, data=catchsavepop, sum)

#average over the last 10 timepoints (keeping split by pop)
catchsavepopmean=aggregate(cbind(catch,cpue)~hrate*pop, data=catchsavepop, mean)

#average over last 10 timepoints (sum of pops)
catchsavetotalmean=aggregate(cbind(catch,cpue)~hrate, data=catchsavetotal, mean)

(fmsy=catchsavetotalmean$hrate[which.max(catchsavetotalmean$catch)])
(Bmsy=catchsavetotalmean$cpue[which.max(catchsavetotalmean$catch)])
```

## Data transformations

The argument `ytrans` in `fitGP_fish` can be used to apply a data transformation to `y` prior to fitting the model. Inputs `y` and `m` should always be in untransformed CPUE. Setting `ytrans="none"` (the default) will apply no tranformation, `ytrans="log"` with compute $log(y)$, `ytrans="gr1"` will compute $log(y_t/m_{t-1})$, and `ytrans="gr2"` will compute $log(y_t/(m_{t-1}-bh_{t-1})$. All $R^2$ values will be computed in the original (untransformed) units of CPUE. Note that the standard deviations in the results table are only for the transformed variable.

The conditional responses (from `getconditionals`) will plot the transformed variable. The `plot` method will plot the untransformed variable (without standard deviations), unless you specify `ytrans=TRUE`. 

```{r ytrans}
fishfit0_trans=fitGP_fish(data=RHlags, y="CPUE_index", m="CPUE_index_1", h="Catch_1", time="Time",
                          ytrans="gr2")
summary(fishfit0_trans)

head(fishfit0_trans$insampresults)

getconditionals(fishfit0_trans)
```

```{r ytransplots, fig.height=3}
plot(fishfit0_trans)
plot(fishfit0_trans, ytrans = TRUE)
```

## Linear prior mean function

There is an option in `fitGP` (and `fitGP_fish`) called `linprior` which will fit a GP model to the residuals of a linear relationship between `y` and the first variable of `x` (prior to scaling) and backtransform as appropriate. Essentially, it fits the model
$$y_t=\beta_0+\beta_1x_{1}+f(x_{1}, x_{2},...)$$
where $f$ is the GP. In other words, the mean function for the GP is assumed to be linear with respect to the first input, rather than constant. The mean function is what the GP reverts to in the absence of data (the larger the inverse length scale is, the more quickly this happens), so altering it can potentially aid in problems with extrapolation. In a fisheries model, if `y` is growth rate (specifically the escapement-based growth rate, `"gr2"`), the first variable of `x` will be the first lag of escapement, and so this linear relationship at acts like a Ricker prior. Option `linprior="local"` fits a separate regression for each population, `"global"` fits a single regression. Defaults to `"none"` (off).

In our example, the fitted function looks largely the same. The difference between this and the previous fit is mainly in what happens when you extrapolate beyond the range of the data. We can see it in the conditional response plot is we set `extrap` to a high number.

```{r}
fishfit0_lin=fitGP_fish(data=RHlags, y="CPUE_index", m="CPUE_index_1", h="Catch_1", time="Time",
                          ytrans="gr2", linprior = "global")
getconditionals(fishfit0_trans, extrap = 1)
getconditionals(fishfit0_lin, extrap = 1)
```

## References

Tsai CH, Munch SB, Masi MD, Stevens MH (2024) Empirical dynamic modelling for sustainable benchmarks of short-lived species. ICES Journal of Marine Science
