---
title: "Variable Timestep Method"
output: rmarkdown::html_vignette
toc: true
vignette: >
  %\VignetteIndexEntry{Variable Timestep Method}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(GPEDM)
library(ggplot2)
```

## Introduction

By default, `fitGP` excludes any rows that contain missing values (NAs) in either the response or predictor variables. Thus, in a delay embedding model, a missing datapoint in the middle of a time series will be excluded as will the following `E` values (assuming `tau` is 1). Stephan Munch and Bethany Johnson developed a method whereby using the time spacing between a value and its lagged predictors as another predictor, we can exclude less data, generate predictions for all non-missing timepoints, and by adjusting the spacing in the forecast matrix, generate forecasts multiple timesteps into the future. Johnson and Munch (2022) contains more details.  

Using this method requires that lag predictors be generated beforehand (option 1 in [Specifying training data](https://tanyalrogers.github.io/GPEDM/index.html#specifying-training-data)). The `makelags` function has an option for the variable timestep method (`vtimestep=TRUE`) that will generate the time difference lags automatically. Including the `time` argument is strongly recommended, and you definitely need to include it if the rows with missing values have already been removed and/or the timesteps are uneven. Argument `Tdiff_max` can be used to set the max time difference value considered (e.g. if you have one large time gap that you don't want to use this method for).

```{r vtimestep, fig.height=3}
data(HastPow3sp)

#add some missing values for variable X
set.seed(20)
HPmiss=HastPow3sp[,c("Time","X")]
HPmiss[sample(1:nrow(HPmiss),100),"X"]=NA
ggplot(HPmiss,aes(x=Time,y=X)) +
  geom_line() +  geom_point() + theme_bw()

#standard method
HPmisslags=makelags(data=HPmiss, y="X", time="Time", E=2, tau=1)
head(cbind(HPmiss,HPmisslags),15)

#variable timestep method
HPmisslags=makelags(data=HPmiss, y="X", time="Time", E=2, tau=1, vtimestep=T)
head(cbind(HPmiss,HPmisslags),15)

HPmissdata=cbind(HPmiss,HPmisslags)
vtdemo=fitGP(data=HPmissdata, y="X", x=colnames(HPmisslags), time="Time")

summary(vtdemo)

basepredplot=ggplot(vtdemo$insampresults,aes(x=timestep,y=predmean)) +
  geom_line() + 
  geom_ribbon(aes(ymin=predmean-predsd,ymax=predmean+predsd), alpha=0.4, color="black") +
  geom_point(aes(y=obs)) +
  theme_bw()
basepredplot
```

## Interpolation

Generate predictions where an observed value is missing.

```{r interp, fig.height=3}
HPmissinterp=HPmissdata[is.na(HPmissdata$X),]
#reinsert true values to get fit statistics
HPmissinterp$X=HastPow3sp$X[is.na(HPmissdata$X)]

vtinterp=predict(vtdemo, newdata = HPmissinterp)
vtinterp$outsampfitstats

basepredplot +
  geom_point(data=vtinterp$outsampresults, aes(y=predmean), color="red") +
  geom_errorbar(data=vtinterp$outsampresults,
                aes(ymin=predmean-predsd,ymax=predmean+predsd),color="red")

```

## Forecasting

You can also generate a forecast matrix using the variable timestep method. The number of timesteps to forecast can be specified with `Tdiff_fore`.

```{r forecast, fig.height=3}
HPmissfore=makelags(data=HPmiss, y="X", time="Time", E=2, tau=1,vtimestep=T,  
                    forecast=T, Tdiff_fore=c(1,2,3))
HPmissfore

vtpred=predict(vtdemo, newdata = HPmissfore)

basepredplot + 
  geom_point(data=vtpred$outsampresults, aes(y=predmean), color="red") +
  geom_errorbar(data=vtpred$outsampresults,
                aes(ymin=predmean-predsd,ymax=predmean+predsd),color="red")
```

## Using augmentation data

When using the variable timestep method, the function `makelags` can also be used to generate an augmentation data matrix that can be passed to `fitGP`. This should work as long as the settings in `makelags` match, except use `augment=TRUE`. When the augmentation table is generated, `makelags` will print a table showing the original number of each Tdiff combination in the original dataset (Freq), and the total number of with the augmentation data included (Freq_new). Combinations are added up to `nreps`, if possible (currently defaults to 10). By default, only Tdiff combinations that appear in the original dataset are used, however, if you supply a vector `Tdiff_fore`, then the augmentation matrix will include or all possible combinations of the Tdiff values supplied in `Tdiff_fore`.

```{r aug, fig.height=3}
HPmisslags=makelags(data=HPmiss, y="X", time="Time", E=2, tau=1, vtimestep=T)
HPaug=makelags(data=HPmiss, y="X", time="Time", E=2, tau=1, vtimestep=T,augment=T)
head(HPaug,15)

HPmissdata=cbind(HPmiss,HPmisslags)
vtdemo_aug=fitGP(data=HPmissdata, y="X", x=colnames(HPmisslags), time="Time",
                 augdata=HPaug)
summary(vtdemo_aug)

vtapred=predict(vtdemo_aug,newdata=HPmissfore)
vtainterp=predict(vtdemo_aug, newdata = HPmissinterp)
vtainterp$outsampfitstats

ggplot(vtdemo_aug$insampresults,aes(x=timestep,y=predmean)) +
  geom_line() + 
  geom_ribbon(aes(ymin=predmean-predsd,ymax=predmean+predsd), alpha=0.4, color="black") +
  geom_point(aes(y=obs)) +
  theme_bw() +  
  geom_point(data=vtapred$outsampresults, aes(y=predmean), color="red") +
  geom_errorbar(data=vtapred$outsampresults,
                aes(ymin=predmean-predsd,ymax=predmean+predsd),color="red")

#vtaloo=predict(vtdemo_aug,predictmethod = "loo")
#vtaseq=predict(vtdemo_aug,predictmethod = "sequential")
#vtloo=predict(vtdemo,predictmethod = "loo")
#vtseq=predict(vtdemo,predictmethod = "sequential")
```

## References

Munch, S. B., Poynor, V., and Arriaza, J. L. 2017. Circumventing structural uncertainty: a Bayesian perspective on nonlinear forecasting for ecology. Ecological Complexity, 32:134.

Johnson, B., and Munch, S. B. 2022. An empirical dynamic modeling framework for missing or irregular samples. Ecological Modelling, 468:109948.
